$$
\newcommand{\Beta}{\textrm{Beta}}
\newcommand{\Bin}{\textrm{Bin}}
$$

## Example 5.1.3 from @BIDA

> Carey et al. (1998) studied the "cost of reproduction" on longevity of Mediterranean fruit flies (medflies). We use data from the 534 medflies that lived at least 34 days. These are cross-classified by two factors: A) whether the fly produced at least 1,000 eggs in its first 30 days of life and B) whether the fly lived at least 44 days, see @tbl-medflies.

| Early egg production | Long-lived | Short-lived |
| --- | ---: | ---: |
| High | 54 | 80 |
| Low | 224 | 176 |
: Medfly longevity by early egg production {#tbl-medflies}

We want to understand the effect of early egg production on medfly longevity. One way to do this is to express the chance of longevity as a function of early egg production
$$
\begin{aligned}
\theta_{H} &\doteq p(\text{Long-lived} \mid \text{High egg production}) \\
\theta_{L} &\doteq p(\text{Long-lived} \mid \text{Low egg production}) 
\end{aligned}
$$
which leads to the model:
$$
y_H \sim \Bin(n_H, \theta_H) \quad \perp \quad y_L \sim \Bin(n_L, \theta_L)
$$
where $n_H = 134$ and $n_L = 400$ is the number of medflies with high and low early egg production, and $y_H = 54$ and $y_L = 224$ is the corresponding number of medflies that lived longer. The sample proportion of long-lived medflies is 0.40 for high early producers and 0.56 for low early producers.

If we assume independent $\operatorname{Beta}\left(a_H, b_H\right)$ and $\operatorname{Beta}\left(a_L, b_L\right)$ priors for $\theta_H$ and $\theta_L$, we obtain the posterior distribution:
$$
\theta_H\left|y_H \sim \operatorname{Beta}\left(y_H+a_H, n_H-y_H+b_H\right) \quad \perp \quad \theta_L\right| y_L \sim \operatorname{Beta}\left(y_L+a_L, n_L-y_L+b_L\right) .
$$


Note that the posterior of $\theta_H$ does not depend on $y_L$ etc. In the absence of available prior information we place independent $U(0,1)=\operatorname{Beta}(1,1)$ reference priors on the probabilities. The posteriors are

$$
\theta_H\left|y_H \sim \operatorname{Beta}(55,81) \quad \perp \quad \theta_L\right| y_L \sim \operatorname{Beta}(225,177).
$$


Our quantites of interest will be.

Relative Risk:
$$
RR = \frac{\theta_L}{\theta_H}
$$

Odds Ratio:
$$
OR = \frac{\theta_L / (1 - \theta_L)}{\theta_H / (1 - \theta_H)}
$$

Risk Difference:
$$
RD = \theta_L - \theta_H
$$

```{r}
#| warning: false
#| messages: false
#| output: false
library(nimble)
library(dplyr)

code <- nimbleCode({
  y[1] ~ dbin(theta[1], n[1])
  y[2] ~ dbin(theta[2], n[2])
  theta[1] ~ dbeta(1, 1)
  theta[2] ~ dbeta(1, 1)
  odds[1] <- theta[1] / (1 - theta[1])
  odds[2] <- theta[2] / (1 - theta[2])
  RD <- theta[2] - theta[1]
  RR <- theta[2] / theta[1]
  OR <- odds[2] / odds[1]
})
```


```{r}
#| warning: false
#| messages: false
#| output: false
constants <- list(n = c(134, 400))
data <- list(y = c(54, 224))
inits <- list(
    list(theta = c(0.1, 0.5)),
    list(theta = c(0.5, 0.1))
)

set.seed(1234)
mcmc.output <- nimbleMCMC(code = code,
                          data = data,
                          constants = constants,
                          inits = inits,
                          monitors = c("theta", "RD", "RR", "OR"),
                          niter = 20000,
                          nburnin = 5000,
                          nchains = 2)
```

```{r}
library(MCMCvis)
MCMCsummary(object = mcmc.output, round = 2)
```

From above, the posterior medians and $95\%$ PIs are  $0.40, (0.32,0.49)$ for the high early producers and $0.56,(0.51,0.61)$ for the low early producers. 

> The interval for the high producers is entirely below the interval for the low producers suggesting that longevity is increased for flies with lower early egg production. 

Similarly, Table 5.2 shows strong evidence that the relative risk and odds ratio are both greater than one and that the risk difference is greater than zero. Medflies with a lower amount of early egg production are estimated to be about 1.38 times more likely to be longlived compared to high early producers and we are $95 \%$ sure that the actual value is in the interval

```{r}
#| warning: false
#| messages: false
#| output: false
# Visualize trace-plot.
MCMCtrace(object = mcmc.output)
```