# Normal model revisited

In this section we will fit a normal distribution with careful choice of prior for the mean and an independent reference prior for the variance. We will work with the following example.

**Example:** Twelve alloy samples were tested using spectrometry by the St. Paul Brass and Aluminum Foundry in St. Paul, Minnesota, in June of 2003. Let $y_i$ be the zinc percentage of sample $i$, $i=1, \ldots, 12$. We assume that the observations are independent and follow a bell-shaped curve. Specifically, we assume they follow a normal distribution with some mean $\mu$ and variance $\sigma^2=1/\tau$ (recall we our use of precision $\tau$). Write

$$y_1, \ldots, y_{12} \mid \mu, \tau \stackrel{i i d}{\sim} N\left(\mu, 1/\tau\right) .$$

The $n=12$ sample zinc percentages are 4.20,4.36,4.11,3.96,5.63,4.50,5.64,4.38,4.45,3.67, 5.26, and 4.66. The sample mean is $\bar{y} .=4.568$, the sample standard deviation is $s=0.631$, so the standard error of $\bar{y}$. is $\operatorname{SE}(\bar{y})=.0.631 / \sqrt{12}=0.182$. The minimum observation is $y_{(1)}=3.67$ and the maximum is $y_{(12)}=5.64$. The first quartile is $q_1=4.133$, the third is $q_3=5.110$, and the median is $\tilde{y}=4.415$.

The data seemes reasonably normal when inspecting a normal plot, and so a normal model is not unreasonable.

```{r}
y <- c(4.20,4.36,4.11,3.96,5.63,4.50,5.64,4.38,4.45,3.67,5.26,4.66)
ybar <- mean(y)
s <- sd(y)
n <- length(y)
qqnorm(y)
qqline(y)
```

See Section 1.2 of @BIDA for more background on this dataset. Based on this data, we want to know what the chance $\Pr(y_{13} < 4.4)$ is for a new observation $y_{13}$. This is the event that zinc will need to be added to the allow.

## Reference prior

As noted in Section 5.2.1 of @BIDA, the most commonly used reference prior $$p(\mu, \tau) \propto \frac{1}{\tau}$$ will give results strikingly similar to frequentist inference for this data. This prior can be thought of as placing "independent" flat priors on both $\mu$ and $\log(\tau)$ (Section 4.6).

See the derivation in the textbook for the form of the joint posterior: $$\begin{aligned}
\mu | \tau, y &\sim N(\bar{y}, 1/(n\tau))\\
\tau | y &\sim \text{Gamma}\left(\frac{n-1}{2}, \frac{(n-1)s^2}{2}\right)
\end{aligned}$$

Next, some elegant distribution theory calculations (see p.g. 11 of @BIDA) that this implies:

$$\frac{\mu - \bar{y}}{s/\sqrt{n}} \Bigg| y \sim t(n-1)\\
\frac{y_{n+1} - \bar{y}}{s\sqrt{1+1/n}} \Bigg| y \sim t(n-1)$$

Note that these results are strikingly similar to frequentist results, but now $\mu$ and $y_{n+1}$ are assumed random (to be drawn from the posterior) while $s$ and $\bar{y}$ are assumed fixed (by conditioning).

This allows us to calculate the chance $\Pr(y_{13} < 4.4) \approx 0.40$ that zinc will be need to be added to a new observation $y_{13}$:

```{r}
pt((4.4 - ybar) / (s * sqrt(1 + 1/n)), df = n - 1)
```

Aside: Read about the fascinating story of ["Student"](../Resources/Students-distr-retrospective.pdf).

## Informative prior

Now let's see what this same analysis says if we use prior information. Following Section 5.2.3 of @BIDA, it is reasonable to assume that our information about $\mu$ and $\tau$ can be independently elicited:
$$p(\mu,\tau) = p(\mu) p(\tau)$$
But this means that we don't have conjugacy anymore, and need to rely on NIMBLE to do our inference.

**Example continued**. We need to identify a prior distribution that gives information about the unknown parameters $\mu$ and $\tau=1 / \sigma^2$. Our expert estimated with $95 \%$ certainty that the mean zinc percentage $\mu$ before pouring should be between $4.5 \%$ and $5 \%$ and would be centered at $4.75 \%$. We interpret this information as $\mathrm{E}(\mu)=4.75$ and $\operatorname{Pr}(4.5<\mu<5)=0.95$ in the prior distribution. Further assuming that the prior on $\mu$ is normal, we find

$$\mu \sim N(4.75,0.0163).$$

(Determining the prior variance of $\mu$ will be discussed shortly.) We have no good information on $\sigma^2$, the variance of an observation $y_i$, so we specify a reference prior on $\sigma^2$ that is independent of $\mu$. The reference prior we use is a gamma distribution on $\tau \equiv 1 / \sigma^2$ with very small parameters. In particular, we used

$$\tau \sim \operatorname{Gamma}(0.001,0.001).$$

Our aim is to learn about $\gamma = \Pr(y_{13} < 4.4) = \Phi((4.4 - \mu) / \sqrt{1/\tau})$. Note that rather than compute the posterior expectation of $\gamma$ like we did previously, we can now also obtain a probability interval.

Here is NIMBLE code for generating the posterior:

```{r}
#| cache: true
library(nimble)
# Model code
zinc_model <- nimbleCode({
  for (i in 1:n) {
    y[i] ~ dnorm(mu, tau)
  }
  mu ~ dnorm(4.75, 1/0.0163)
  tau ~ dgamma(0.001, 0.001) # dflat()
  sigma <- 1 / sqrt(tau)
  gamma_prob <- phi((4.4 - mu) / sqrt(1 / tau))
  prob <- step(4.4 - y[13])
})

# Data and constants
zinc_data <- list(
  y = c(4.20, 4.36, 4.11, 3.96, 5.63, 4.50,
        5.64, 4.38, 4.45, 3.67, 5.26, 4.66, NA)
)

zinc_constants <- list(n = 13)

# Initial values
zinc_inits <- list(mu = 0, tau = 1)

# Build and compile model
zinc_mcmc <- nimbleMCMC(
  code      = zinc_model,
  data      = zinc_data,
  inits     = zinc_inits,
  constants = zinc_constants,
  monitors  = c("mu", "tau", "sigma", "gamma_prob", "prob"),
  niter     = 50000,
  nburnin   = 2000,
  nchains   = 3,
  summary   = TRUE
)
```

```{r}
round(zinc_mcmc$summary$all.chains, 3)
```

Note that $\Pr(y_{13} < 4.4|y) \approx 0.33$. In fact, we can study

$$\gamma = \Pr(y_{13} < 4.4|y, \mu, \tau) = \Phi\left(\frac{4.4-\mu}{\sqrt{1/\tau}}\right).$$

We find that $E[\gamma|y] \approx 0.33$ while $\gamma \in (0.20, 0.45)$ with 95% chance.

## Eliciting the prior

In order to illustrate our prior choice, let us consider
$$
y_1, \ldots, y_n \sim N(\mu, 1/\tau) 
$$
representing final scores of $n$ students in my STA 345 class that I teach every year. (These are typically multi-modal and thus not normal but we will ignore that.)

We will assume prior independence between $\mu$ and $\tau$ (or $\sigma^2 = \frac{1}{\tau}$) and consider a prior that has the form
$$
\begin{aligned}
\mu &\sim N(a, 1/b)\\
\tau &\sim \textrm{Gamma}(c,d) \quad \text{ or } \quad  \sigma^2 \sim \textrm{Gamma}(c,d)
\end{aligned}
$$

So there are four parameters to elicit, so I need to ask (myself) at least four questions to elicit these parameters. 

### Elicit prior on $\mu$

$\mu$ is the average exam score in my class. I can use previous year's information about these topics.

> My best guess for $\mu$ is 65.

So I take my point estimate for $\mu$ to be equal to $a = 65$ 

> I am virtually certain that $\mu$ will not exceed $u = 70$.

So I choose $b$ so that:
$$
\Pr(\mu \leq 70) = 0.95
$$
or equivalently
$$
a + 1.645 \sqrt{\frac{1}{b}} = 0.7
$$
This gives $b = 1.108$, and thus
$$
\mu \sim N(65, \frac{1}{1.108})
$$

### Elicit prior on $\tau$

Elicit prior on variance parameters directly is hard. The idea is relate them to quantiles of the observations $y_1, \ldots, y_n$, which are easier to reason about.

Let $\gamma_{0.9}$ be the 90th percentile score for my class. Why does this help to elicit prior on $\tau$ (or $\sigma$)? Because based on our normal model:
$$
\gamma_{0.9} = \mu + 1.28 \sigma
$$

> My guess for the 90th percentile score in my class is $\tilde{\gamma}_{0.9} = 85$.

We set this equal to guess:
$$
a + 1.28 \sigma_0 = \tilde{\gamma}_{0.9} = 85
$$
which shows
$$
\sigma_0 = (\tilde{\gamma}_{0.9} - a) / 1.28 = 15.625
$$
or 
$$
\tau_0 = 1 / \sigma_0^2 = 0.004096.
$$


We can set this guess as the mode for our Gamma distribution on $\tau$ or $\sigma$, but we need one more piece of information.

> I am virtually certain that the 90th percentile score in my class will not exceed $\tilde{u} = 91$.

We take this to mean that

$$
\begin{aligned}
0.95 &= \Pr(\mu + z_{0.9} \sigma \leq \tilde{u} | \mu = a)\\
&= \Pr\left(\sigma \leq \frac{\tilde{u} - a}{z_{0.9}} \right)\\
&= \Pr\left(\tau \geq \frac{z_{0.9}^2}{(\tilde{u} - a)^2}\right)
\end{aligned}.  
$$

Thus this provides us information on the 95th percentile of $\sigma$ or equivalently the 5th percentile of $\tau$. 
$$
\tilde{\sigma}_{0.95} = \frac{\tilde{u} - a}{z_{0.9}} = 20.3125 \quad \text{ or } \quad \tilde{\tau}_{0.05} = \frac{z_{0.9}^2}{(\tilde{u} - a)^2} = 0.002048.
$$
We can use this information to find the parameters of our Gamma distribution on $\tau$ (or $\sigma$) by trial and error.

First assuming $\sigma \sim \textrm{Gamma}(e,f)$, we equate the mode to $\sigma_0$:
$$
\sigma_0 = \frac{e-1}{f} \quad \text{ or } \quad e = 1 + \sigma_0 f
$$

Now we need to find $f$ such that the 95th percentile of $\sigma$ is $\tilde{\sigma}_{0.95} = 20.3125$. We find that
$$
\sigma \sim \textrm{Gamma}(41.4375,2.588)

```{r}
alpha <- 0.90 
beta <- 0.95 
a <- 65 # Best guess for mu 
tildegamma <- 85 # Best guess for gamma_alpha
tildeu <- 91 # Best guess percentile of gamma_alpha 
zalpha <- 1.28 # qnorm(0.90,0,1) 
f <- 3 # Initial value for f 
# Could use a sequence of values, say f <- seq(1,50,1) 
sigma0 <- (tildegamma - a)/zalpha 
e <- 1 + sigma0*f 
# We must find the Gamma(e,f) distribution that 
# has beta-percentile = tildesigmabeta 
tildesigmabeta <- (tildeu - a)/zalpha 
trialq <- qgamma(beta,e,f) # Return beta-percentile for the 
                           # selected gamma distribution  
trialq       # If trialq = tildesigmabeta
tildesigmabeta # stop and pick corresponding f
```