```{r setup}
#| include: false
library(nimble)
library(tidyverse)
library(knitr)
```

The following analysis is taken from Section 5.2.5 of @BIDA.

## Motivating Application: Renal Osteodystrophy {#sec-application}

> **Clinical context.** Renal osteodystrophy is a bone disease arising when failing kidneys cannot maintain proper calcium and phosphorus levels. Distinguishing **low** from **normal** bone turnover guides disease management.

The commercially available **Diasorin** assay claims to separate these two patient groups. A cross-section of **34 kidney patients** from the bone registry at the University of Kentucky were classified as low or normal turnover by independent means and then given the Diasorin assay. The research question:

> *Can the Diasorin score reliably classify a patient as low vs. normal bone turnover?*

### The Data

```{r data}
low    <- c(91, 46, 95, 60, 33, 410, 105, 43, 189, 1097,
            54, 178, 114, 137, 233, 101, 25, 70, 357)
normal <- c(370, 267, 99, 157, 75, 1281, 48, 298,
            268, 62, 804, 430, 171, 694, 404)

n1 <- length(low) # low group
n2 <- length(normal) # normal group

cat(c(Low=n1,Normal=n2))
```

```{r}
#| echo: false
df_raw <- data.frame(
  score = c(low, normal),
  group = factor(rep(c("Low", "Normal"), c(n1, n2)))
)

df_log <- df_raw |> mutate(log_score = log(score))
```

The low group has `r n1` patients and the normal group has `r n2` patients.

### Data Visualization

```{r fig-5-2a}
#| fig-cap: "**Figure 5.2a.** Boxplots on original scale."
#| echo: false
ggplot(df_raw, aes(group, score, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 1) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Original scale", x = NULL, y = "Diasorin") +
  theme_bw(base_size = 12) + theme(legend.position = "none")
```

```{r fig-5-2b}
#| fig-cap: "**Figure 5.2b.** Boxplots on log scale."
#| echo: false

ggplot(df_log, aes(group, log_score, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 1) +
  scale_fill_manual(values = c("steelblue", "tomato")) +
  labs(title = "Log scale", x = NULL, y = "log(Diasorin)") +
  theme_bw(base_size = 12) + theme(legend.position = "none")
```

```{r fig-5-2c}
#| fig-cap: "Normal Q–Q plots (log scale)."

ggplot(df_log, aes(sample = log_score)) + 
  stat_qq(color = "steelblue") + stat_qq_line() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
        facet_wrap(~group) + 
  theme_bw(base_size = 12)
```

While the original data is right skewed, after log transformation it appears to follow a normal model.

## The Statistical Model {#sec-model}

### Sampling Model

We assume independent samples from two normal populations on the **log scale**:

$$y_{11}, \ldots, y_{1n_1} \mid \mu_1, \tau_1 \overset{\text{iid}}{\sim} N(\mu_1, 1/\tau_1)$$

$$y_{21}, \ldots, y_{2n_2} \mid \mu_2, \tau_2 \overset{\text{iid}}{\sim} N(\mu_2, 1/\tau_2)$$

The two groups are assumed independent.

### Allowing Unequal Variances

Unlike frequentist approaches, Bayesian inference has **no technical difficulty** with unequal variances: we simply assign separate priors to $\tau_1$ and $\tau_2$.

However, unequal variances create an **interpretive challenge**:

> If $\mu_2 > \mu_1$ but $\sigma_1 \gg \sigma_2$, a higher mean for population 2 does **not** imply that a randomly chosen individual from population 2 scores higher.

```{r fig-5-3}
#| fig-cap: "**Figure 5.3.** Two normal distributions with different means and variances. Population 1 has a much higher proportion above 7, even though its mean is lower."
#| fig-height: 4
x_seq <- seq(0, 10, length.out = 500)
tibble(
  x   = rep(x_seq, 2),
  y   = c(dnorm(x_seq, 5, 2), dnorm(x_seq, 5.5, 0.4)),
  Pop = rep(c("Population 1: N(5, 2²)", "Population 2: N(5.5, 0.4²)"), each = 500)
) |>
  ggplot(aes(x, y, color = Pop, linetype = Pop)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 7, linetype = "dotted", color = "gray40") +
  annotate("text", x = 7.15, y = 0.9, label = "cutoff = 7", hjust = 0, size = 3.5) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(x = NULL, y = "Density", color = NULL, linetype = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

**Takeaway:** When variances differ substantially, comparing means alone is misleading. The textbook recommends **examining predictive distributions** rather than just posterior means.

## Prior Elicitation {#sec-priors}

The prior was elicited by asking Dr. Johann Herberth MD, PhD, MPH (Division of Nephrology, Bone and Mineral Metabolism; University of Kentucky).

### Eliciting priors on $\mu_1$ and $\mu_2$.

Because we work on the **log scale**, we elicit beliefs about the **median** of the original Diasorin values. If $y \sim N(\mu, 1/\tau)$ on the log scale, then the median of the original values is $e^\mu$.

The procedure we use will be as follows:

- Expert specifies best-guess median $\tilde{m}$ and an upper bound $\tilde{u}$ such that $P(\text{median} \leq \tilde{u}) = 0.95$.
- Since $P(\mu \leq \log \tilde{m}) = 0.50$ and $P(\mu \leq \log \tilde{u}) = 0.95$, for the prior $\mu \sim N(a, b)$ we have:

$$a = \log(\tilde{m}), \qquad b = \frac{[\log(\tilde{u}) - \log(\tilde{m})]^2}{1.645^2}$$

Dr. Herberth provided the guess of the median Diasorin values in the low bone tunrover group ($n_1=19$) to be $\tilde{m}_L = 130$, and he was 95% sure that this median was less than $\tilde{u}_L = 142$. For the normal bone tunrover group ($n_2=15$), their guess was $\tilde{m}_N = 220$ and $\tilde{u}_N = 240$.

This leads to the prior choices:

```{r prior-mu}
#| echo: false
m_tilde_L <- 130;  u_tilde_L <- 142
a_L <- log(m_tilde_L)
b_L <- (log(u_tilde_L) - log(m_tilde_L))^2 / 1.645^2
m_tilde_N <- 220;  u_tilde_N <- 240
a_N <- log(m_tilde_N)
b_N <- (log(u_tilde_N) - log(m_tilde_N))^2 / 1.645^2
cat(sprintf("Low group:    mu_L ~ N(%.4f, %.5f)\n", a_L, b_L))
cat(sprintf("Normal group: mu_N ~ N(%.4f, %.5f)\n", a_N, b_N))
```

### Eliciting priors on $\tau_1$ and $\tau_2$

We elicit the expert's best guess for the **90th percentile** of Diasorin values in each group. Conditional on the best-guess median, if the 90th percentile guess is $\tilde{\gamma}_{0.9}$, then we have

$$\tau_0 = \frac{1.645^2}{\left[\log(\tilde{\gamma}_{0.9}/\tilde{m})\right]^2}$$

We then set $c = 1 + \tau_0 d$ for a $\text{Gamma}(c, d)$ prior on $\tau$, where $d$ is chosen to reflect our uncertainty (small $d$ = vague, larger $d$ = more informative; see [visualization](https://distribution-explorer.github.io/continuous/gamma.html)).

Dr. Herberth provided his best guess for the 90th percentile in the low and normal bone tunrover group to be $\tilde{\gamma}_{0.9, L} = 170$ and $\tilde{\gamma}_{0.9, N} = 280$, respectively. The meant that our median of our Gamma distributions on $\tau$ for the two groups was:

```{r prior-tau}
#| echo: false
# Expert guess for 90th percentile
g90_L <- 170;  g90_N <- 280

tau0_L <- 1.645^2 / (log(g90_L / m_tilde_L))^2
tau0_N <- 1.645^2 / (log(g90_N / m_tilde_N))^2
cat(sprintf("tau0 (low):    %.4f\n", tau0_L))
cat(sprintf("tau0 (normal): %.4f\n", tau0_N))
```

For the vague choice $d=0.001$, this results in the priors:

```{r}
#| echo: false
d_val  <- 0.001
c_L    <- 1 + tau0_L * d_val
c_N    <- 1 + tau0_N * d_val
cat(sprintf("tau_L ~ Gamma(%.5f, %.3f)\n", c_L, d_val))
cat(sprintf("tau_N ~ Gamma(%.5f, %.3f)\n", c_N, d_val))
```

::: callout-warning
#### Sensitivity of tau priors

With $d = 0.001$, the Gamma means are on the order of $\frac{c}{d}$ 1000 — placing 97% of prior probability **above** the mode. This is a weakly informative prior that strongly favors large precision values. The textbook performs a **sensitivity analysis** varying $d \in \{0.001, 0.01, 0.1\}$ and also varying the prior over $\mu$. (Exercise)
:::

------------------------------------------------------------------------

## NIMBLE Implementation {#sec-nimble}

### Model Code

```{r nimble-model}
diasorin_code <- nimbleCode({
  # Likelihood (log-transformed data)
  for (i in 1:n1) { y1[i] ~ dnorm(mu[1], tau = tau[1]) }
  for (j in 1:n2) { y2[j] ~ dnorm(mu[2], tau = tau[2]) }

  # Priors (informative)
  mu[1]  ~ dnorm(a[1], sd = sqrt(b[1]))
  mu[2]  ~ dnorm(a[2], sd = sqrt(b[2]))
  tau[1] ~ dgamma(c[1], rate = d[1])
  tau[2] ~ dgamma(c[2], rate = d[2])

  # Derived quantities
  sigma[1] <- 1 / sqrt(tau[1])
  sigma[2] <- 1 / sqrt(tau[2])
  delta    <- mu[2] - mu[1]          # Difference in log-scale means
  tau_ratio <- tau[2] / tau[1]

  # Predictive draws
  yPred1 ~ dnorm(mu[1], tau = tau[1])
  yPred2 ~ dnorm(mu[2], tau = tau[2])

  # Back-transform to original scale
         # Relative median (normal / low)
  predO1 <- exp(yPred1)              # Predictive original scale, low
  predO2 <- exp(yPred2)              # Predictive original scale, normal

  # Posterior probability that normal > low
  P_mu   <- step(delta)
})
```

### Data and Initial Values

```{r nimble-data}
log_low    <- log(low)
log_normal <- log(normal)

nimble_data <- list(
  y1 = log_low,
  y2 = log_normal
)

nimble_constants <- list(
  n1 = n1, n2 = n2,
  a  = c(a_L,  a_N),
  b  = c(b_L,  b_N),   # precision for mu prior
  c  = c(c_L,  c_N),
  d  = c(d_val, d_val)
)

nimble_inits <- list(
  mu    = c(mean(log_low), mean(log_normal)),
  tau   = c(1, 1)
)
```

### Running MCMC

```{r run-mcmc}
#| results: hide
#| cache: true
set.seed(2024)

model <- nimbleModel(
  code      = diasorin_code,
  data      = nimble_data,
  constants = nimble_constants,
  inits     = nimble_inits
)

compiled_model <- compileNimble(model)

mcmc_conf <- configureMCMC(model,
  monitors = c("mu", "tau", "sigma", "delta", "tau_ratio",
               "yPred1", "yPred2",
               "predO1", "predO2", "P_mu"))

mcmc      <- buildMCMC(mcmc_conf)
c_mcmc    <- compileNimble(mcmc, project = model)

c_mcmc$run(50000)
samples   <- as.matrix(c_mcmc$mvSamples)

# Discard burn-in
burn_in   <- 5000
samples   <- samples[(burn_in + 1):nrow(samples), ]
```

## Posterior Results {#sec-results}

### Summary Table (Table 5.3 Equivalent)

```{r table-results}
post_summary <- function(samp, vars) {
  sapply(vars, function(v) {
    x <- samp[, v]
    c(mean = mean(x), sd = sd(x),
      `2.5%` = quantile(x, 0.025),
      `50%`  = median(x),
      `97.5%` = quantile(x, 0.975))
  }) |> t() |> as.data.frame() |>
    round(3)
}

vars_log <- c("mu[1]", "mu[2]", "delta", "tau[1]", "tau[2]", "tau_ratio")
tbl_log  <- post_summary(samples, vars_log)
rownames(tbl_log) <- c("µ₁ (Low)", "µ₂ (Normal)", "µ₂ − µ₁", "τ₁", "τ₂", "τ₂/τ₁")

kable(tbl_log, caption = "Table 5.3. Posterior summaries (log scale, informative prior).",
      digits = 3, align = "r")
```

::: callout-important
#### Interpretation

- $P(\mu_2 > \mu_1 \mid y) \approx 1$ — the normal group has a decisively higher log-scale mean.
- The **predictive probability** that a new normal-group observation exceeds a new low-group observation is only about **0.66**
:::

```{r}
mean(samples[,"yPred2"] > samples[,"yPred1"])
```

## Predictive Distributions {#sec-predictive}

### Predictive Densities — Log Scale (Figure 5.4)

```{r fig-5-4}
#| fig-cap: "**Figure 5.4.** Posterior predictive densities on the log scale."
#| fig-height: 4

cutoff <- (median(samples[,"mu[1]"]) + median(samples[,"mu[2]"]))/2

tibble(
  value = c(samples[, "yPred1"], samples[, "yPred2"]),
  group = rep(c("Low", "Normal"), each = nrow(samples))
) |>
  ggplot(aes(value, fill = group, color = group)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_fill_manual(values  = c("steelblue", "tomato")) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  labs(x = "log(Diasorin Score)", y = "Predictive Density",
       fill = NULL, color = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top") +
  geom_vline(xintercept=cutoff)
```

Misclassification probabilities:

```{r}
mean(samples[, "yPred1"] > cutoff)
mean(samples[, "yPred2"] < cutoff)
```

### Predictive Densities — Original Scale (Figure 5.5)

```{r fig-5-5}
#| fig-cap: "**Figure 5.5.** Posterior predictive densities on the original scale. Note the severe right skew — means and SDs are uninformative here."
#| fig-height: 4
tibble(
  value = c(samples[, "predO1"], samples[, "predO2"]),
  group = rep(c("Low", "Normal"), each = nrow(samples))
) |>
  ggplot(aes(value, fill = group, color = group)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_fill_manual(values  = c("steelblue", "tomato")) +
  scale_color_manual(values = c("steelblue", "tomato")) +
  coord_cartesian(xlim = c(0, 1500)) +
  labs(x = "Diasorin Score", y = "Predictive Density",
       fill = NULL, color = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top") +
  geom_vline(xintercept=cutoff)
```

::: callout-note
The log-scale predictive distributions are much easier to compare. The original-scale densities are **log-normal**: highly skewed, overlapping substantially. The textbook rightly emphasizes that **plotting predictive distributions** — not just reporting posterior means — is essential for sound conclusions when variances differ.
:::