The analysis here follows Section 9.4 of @BIDA.

```{r setup, include=FALSE}
library(Hmisc)
library(nimble)
library(ggplot2)
library(coda)
```

# Motivation: The FEV Study

The **Forced Expiratory Volume (FEV)** dataset (Rosner, 2006) is introduced as Example 7.1.1 in @BIDA, which measures pulmonary function in adolescents. The response $y$ is the volume of air (liters) expelled in 1 second of a forceful breath.

Lung function is expected to increase during adolescence, but smoking may slow its progression. We have $n = 345$ adolescents aged 10–19. The predictor variables include age and smoking status (NS-not current smoker and S-current smoker).

**Scientific question:** Does smoking slow the development of lung capacity during adolescence?

## The Confounding Problem

First we obtain and filter the dataset from the `Hmisc` package:

```{r load-data}
library(Hmisc)
getHdata(FEV)

fev <- subset(FEV, age >= 10 & age <= 19)   # 345 adolescents as in textbook
# create a binary variable for smoking status
fev$smoke01  <- as.numeric(fev$smoke == "current smoker")
# we create an interaction term for regression
fev$ageSmoke <- fev$age * fev$smoke01
```

Here are some details of the dataset:

```{r}
#| echo: false
cat("Dataset dimensions:", nrow(fev), "x", ncol(fev), "\n")
cat("Age range:", range(fev$age), "\n")
cat("Smokers:", sum(fev$smoke01), "  Nonsmokers:", sum(1 - fev$smoke01), "\n")
cat("Mean FEV:", mean(fev$fev[fev$smoke01 == 1]), " (smokers) vs ", mean(fev$fev[fev$smoke01 == 0]), " (nonsmokers)\n")
```

While the last metric suggests smokers have better lung function than non-smokers, this is not the case. The problem is confounding due to age: older adolescents are more likely to smoke and have higher lung function!

We need to account for `age` in our analysis. The following plot suggest that smoking indeed degrades FEV.

```{r confounding-plot, fig.height=4, fig.width=7}
#| echo: false
#| warnings: false
#| message: false
ggplot(fev, aes(x = age, y = fev, color = smoke, shape = smoke)) +
  geom_jitter(size = 1.5, width = 0.25, height=0, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(
    title = "FEV vs Age by Smoking Status",
    subtitle = "Confounding is visible: smokers tend to be older",
    x = "Age in years (jittered)", y = "FEV (liters)",
    color = "Smoking Status", shape = "Smoking Status"
  ) +
  theme_bw()
```

## The Regression Model

We fit a normal linear model with an **interaction between age and smoking status**:

$$y_i \mid \beta, \tau \sim N(x_i'\beta,\; \tau^{-1}), \quad i = 1, \ldots, n$$

with predictor vector $x_i' = (1,\; \text{age}_i,\; \text{smoke}_i,\; \text{age}_i \cdot \text{smoke}_i)$ and parameter vector $\beta = (\beta_1, \beta_2, \beta_3, \beta_4)'$.

Thus for non-smokers ($\text{smoke}_i = 0$) we fit the model:
$$y_i = \beta_1 + \beta_2 \text{age}_i + \epsilon_i,$$
while for smokers ($\text{smoke}_i = 1$) we fit
$$y_i = (\beta_1 + \beta_3) + (\beta_2 + \beta_4) \text{age}_i + \epsilon_i$$

**Interpretation of coefficients:**

- $\beta_1$: intercept (baseline FEV at "age 0" for nonsmokers)
- $\beta_2$: rate of FEV increase per year for *nonsmokers*
- $\beta_3$: difference in intercepts between smokers and nonsmokers (at "age 0")
- $\beta_4$: difference in FEV increase per year between smokers and nonsmokers (the key parameter for our question)

```{r design-matrix}
X <- cbind(1, fev$age, fev$smoke01, fev$ageSmoke)
y <- fev$fev
colnames(X) <- c("Intercept", "Age", "Smoke", "Age:Smoke")
X[sample(nrow(X), 10),]
```

## Bayesian analysis

For a Bayesian analysis, we need to assume priors a $p(\beta, \tau)$. Here we will focus on informative priors, but for your last homework you will compare the results with SIR priors.

Following Section 9.4 of @BIDA, it seems easiest to elicit prior when we assume prior independence of $\beta$ and $\tau$:

$$p(\beta, \tau) = p(\beta) \cdot p(\tau)$$

with

$$\beta \sim N_r(\beta_0, C_0) \quad \perp\!\!\!\perp \quad \tau \sim \mathrm{Gamma}(a, b).$$

Our task is to determine $\beta_0$, $C_0$, $a$, and $b$ by asking our expert questions about their beliefs. The key insight of the **BCJ (Branscum–Christensen–Johnson) method** is to elicit priors on *observable conditional means* rather than on $\beta$ directly.

------------------------------------------------------------------------

# Prior Elicitation for $\beta$ 

## The BCJ Strategy

Individual regression coefficients $\beta_j$ are abstract — they have no direct physical interpretation. Instead, the BCJ method elicits priors on conditional means

$$\tilde{m}_i \equiv E[y \mid \tilde{x}_i] = \tilde{x}_i'\beta$$

for $r = 4$ subpopulations defined by covariate vectors $\tilde{x}_1, \ldots, \tilde{x}_4$ that an expert can assess **independently**. We then *induce* a prior on $\beta$.

## Choosing the Elicitation Points

We choose four covariate combinations that span the age and smoking space:

| Group | $\tilde{x}_i'$   | Description           |
|-------|------------------|-----------------------|
| 1     | $(1, 11, 0, 0)$  | 11-year-old nonsmoker |
| 2     | $(1, 13, 1, 13)$ | 13-year-old smoker    |
| 3     | $(1, 16, 0, 0)$  | 16-year-old nonsmoker |
| 4     | $(1, 18, 1, 18)$ | 18-year-old smoker    |

This gives the **design matrix for elicitation**:

$$\tilde{X} = \begin{bmatrix} 1 & 11 & 0 & 0 \\ 1 & 13 & 1 & 13 \\ 1 & 16 & 0 & 0 \\ 1 & 18 & 1 & 18 \end{bmatrix}$$

```{r Xtilde}
X_tilde <- matrix(c(1, 11,  0,  0,
                    1, 13,  1, 13,
                    1, 16,  0,  0,
                    1, 18,  1, 18),
                  nrow = 4, byrow = TRUE)
rownames(X_tilde) <- c("11yr NS", "13yr S", "16yr NS", "18yr S")
colnames(X_tilde) <- c("1", "Age", "Smoke", "Age:Smoke")
X_tilde
```

## Eliciting Normal Priors on Conditional Means

A collaborator with domain knowledge (a pulmonologist, say) provides the following assessments. For each group, we elicit a best guess $\tilde{y}_i$ (prior mean of $\tilde{m}_i$) and a bound that translates to a prior variance $\tilde{w}_i$.

**Example: 18-year-old smokers** ($\tilde{m}_4$). The expert believes:

- Best guess for mean FEV: $\tilde{y}_4 = 3.3$ liters
- 99% sure the mean FEV is less than 4.0 liters

Under a normal prior, the 99th percentile satisfies $4.0 = 3.3 + 2.33\sqrt{\tilde{w}_4}$, so

$$\tilde{w}_4 = \frac{(4.0 - 3.3)^2}{2.33^2} = \frac{0.49}{5.4289} \approx 0.09$$

Applying the same logic to all four groups gives:

$$\tilde{m} = \tilde{X}\beta \sim N_4\!\left( \begin{bmatrix} 2.8 \\ 3.0 \\ 4.0 \\ 3.3 \end{bmatrix},\; \begin{bmatrix} 0.04 & 0 & 0 & 0 \\ 0 & 0.04 & 0 & 0 \\ 0 & 0 & 0.04 & 0 \\ 0 & 0 & 0 & 0.09 \end{bmatrix} \right)$$

```{r elicitation}
y_tilde <- c(2.8, 3.0, 4.0, 3.3)
D_tilde <- diag(c(0.04, 0.04, 0.04, 0.09))
```

## Inducing the Prior on $\beta$

Since $\tilde{m} = \tilde{X}\beta$ is multivariate normal and $\tilde{X}$ is invertible, we have $\beta = \tilde{X}^{-1}\tilde{m}$, which gives

$$\beta \sim N_4\!\left(\tilde{X}^{-1}\tilde{Y},\; \tilde{X}^{-1} D(\tilde{w}) \tilde{X}^{-1\prime}\right)$$

```{r induced-prior-beta}
X_tilde_inv <- solve(X_tilde)

beta0 <- X_tilde_inv %*% y_tilde
C0    <- X_tilde_inv %*% D_tilde %*% t(X_tilde_inv)

cat("Prior mean for beta:\n")
print(round(beta0, 4))

cat("\nPrior covariance matrix for beta:\n")
print(round(C0, 4))

cat("\nPrior standard deviations for beta:\n")
print(round(sqrt(diag(C0)), 4))
```

**Note:** The prior on $\beta$ is *not* diagonal — the components of $\beta$ are correlated a priori, even though the $\tilde{m}_i$ were specified independently. This is an automatic consequence of the linear relationship $\tilde{m} = \tilde{X}\beta$.

# Prior Elicitation for $\tau$

## Mode-and-Percentile Elicitation

To choose $a$ and $b$ in $\tau \sim \mathrm{Gamma}(a, b)$, we elicit two pieces of information about **individual-level variability** (not about the mean FEV).

**Step 1 — Best guess for the 95th percentile of FEV among 18-year-old smokers.**

Under normality, the 95th percentile of FEV is $\tilde{m}_4 + 1.645\sigma$. With $\tilde{m}_4 = 3.3$ (our best guess) and the **expert's best guess for the 95th percentile being 5.0**, we get:

$$5.0 = 3.3 + 1.645\sigma_0 \implies \sigma_0 = \frac{1.7}{1.645} \implies \tau_0 = \left(\frac{1.645}{1.7}\right)^2 \approx 0.94$$

This gives the **mode** of the Gamma prior: $\frac{a-1}{b} = 0.94$, i.e. $a = 0.94b + 1$.

**Step 2 — Lower bound on the 95th percentile.**

Suppose further that the expert is 95% sure that the 95th percentile of FEV for 18-year-old smokers exceeds 4.0:

$$0.95 = P(\tilde{m}_4 + 1.645\sigma > 4.0| \tilde{m}_4 = 3.3) = P\!\left(\tau < \left(\frac{1.645}{0.7}\right)^2\right) = P(\tau < 5.52)$$

We find $b$ such that $\mathrm{Gamma}(0.94b+1,\, b)$ has its 95th percentile equal to 5.52.

```{r tau-prior}
b_grid <- seq(0.01, 1.5, by = 0.01)
a_grid <- 0.94 * b_grid + 1
p95    <- pgamma(5.52, shape = a_grid, rate = b_grid)

# Find b where p95 is closest to 0.95
idx <- which.min(abs(p95 - 0.95))

cat("Approximate solution:\n")
cat(sprintf("  b = %.2f,  a = %.4f,  P(tau < 5.52) = %.5f\n",
            b_grid[idx], a_grid[idx], p95[idx]))

# Show nearby values (like the textbook table)
print(data.frame(a = round(a_grid[(idx-2):(idx+2)], 4),
                 b = b_grid[(idx-2):(idx+2)],
                 p95 = round(p95[(idx-2):(idx+2)], 5)))

a_tau <- a_grid[idx]
b_tau <- b_grid[idx]
```

# Posterior Inference with NIMBLE

## NIMBLE Model Specification

```{r nimble-model, results='hide'}
fev_code <- nimbleCode({
  # Likelihood
  for (i in 1:n) {
    y[i] ~ dnorm(mu[i], tau = tau)
    mu[i] <- inprod(beta[1:r], X[i, 1:r])
  }

  # Independence prior: beta independent of tau
  beta[1:r] ~ dmnorm(beta0[1:r], cov = C0[1:r, 1:r])
  tau        ~ dgamma(a_tau, b_tau)

  # Posterior predictive for a new 18-year-old smoker
  mu_new <- inprod(beta[1:r], x_new[1:r])
  y_new  ~ dnorm(mu_new, tau = tau)

  # Posterior predictive for a new 18-year-old nonsmoker
  mu_new2 <- inprod(beta[1:r], x_new2[1:r])
  y_new2 ~ dnorm(mu_new2, tau = tau)
})
```

## Data and Constants

```{r nimble-data}
r <- 4

x_new <- c(1, 18, 1, 18)   # 18-year-old smoker
x_new2 <- c(1, 18, 0, 0)   # 18-year-old nonsmoker

fev_data <- list(
  y = y,
  X = X
)

fev_const <- list(
  n     = nrow(X),
  r     = r,
  beta0 = as.vector(beta0),
  C0    = C0,
  a_tau = a_tau,
  b_tau = b_tau,
  x_new = x_new,
  x_new2 = x_new2
)

fev_inits <- list(
  beta  = as.vector(beta0),
  tau   = 1.0,
  y_new = 3.3,
  y_new2 = 3.3
)
```

## Running the MCMC

```{r nimble-run, cache=TRUE, results='hide'}
set.seed(577)

samples  <- nimbleMCMC(code=fev_code, constants = fev_const,
                          data = fev_data, inits = fev_inits,
                          nchains = 2, 
                          #niter = 30000, nburnin = 1000,
                          niter = 100000, nburnin = 20000,
                          monitors = c("beta", "tau", "mu_new", "y_new", "y_new2"),
                          samplesAsCodaMCMC = TRUE)
```

## Convergence Diagnostics

```{r convergence, fig.height=5, fig.width=8}
library(MCMCvis)
MCMCsummary(samples, round = 2)
```

```{r traceplots}
# I had to check this and increase niter and nburnin
MCMCtrace(samples)
```

## Posterior Summaries

```{r posterior-summary}
all_samples <- do.call(rbind, samples)

params <- c("beta[1]","beta[2]","beta[3]","beta[4]","tau","mu_new","y_new", "y_new2")
summ <- t(apply(all_samples[, params], 2, function(x)
  c(Mean = mean(x), SD = sd(x),
    `2.5%` = unname(quantile(x, 0.025)),
    `97.5%` = unname(quantile(x, 0.975))
    )))

knitr::kable(round(summ, 4))
```

## Interpretation

```{r interpretation}
beta_samples <- all_samples[, paste0("beta[", 1:4, "]")]

# P(beta4 < 0): is the smoking effect on the age slope negative?
p_neg_slope <- mean(beta_samples[, 4] < 0)
cat(sprintf("P(beta4 < 0 | data) = %.3f\n", p_neg_slope))

# Estimated smoking effect at age 15
smoke_effect_15 <- beta_samples[, 3] + 15 * beta_samples[, 4]
cat(sprintf("\nSmoking effect on mean FEV at age 15:\n"))
cat(sprintf("  Mean: %.3f  SD: %.3f  95%% CI: (%.3f, %.3f)\n",
            mean(smoke_effect_15), sd(smoke_effect_15),
            quantile(smoke_effect_15, 0.025),
            quantile(smoke_effect_15, 0.975)))
```

```{r posterior-pred-plot, fig.height=4, fig.width=7}
# Posterior predictive distribution for an 18-year-old smoker
y_new_samples <- all_samples[, "y_new"]
y_new2_samples <- all_samples[, "y_new2"]

# A dodged historgram of y_new and y_new2 samples 
plot_df <- data.frame(
  fev = c(y_new_samples, y_new2_samples),
  group = factor(
    rep(c("Smoker", "Nonsmoker"), c(length(y_new_samples), length(y_new2_samples))),
    levels = c("Smoker", "Nonsmoker")
  )
)

ggplot(plot_df, aes(x = fev, fill = group)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, alpha = 0.45,
                 position = "identity", color = "white") +
  labs(
    title = "Posterior Predictive FEV: 18-year-old Smoker vs Nonsmoker",
    x = "FEV (liters)",
    y = "Density") +
  scale_fill_manual(values = c("Smoker" = "coral", "Nonsmoker" = "lightblue")) +
  scale_color_manual(values = c("Smoker" = "navy", "Nonsmoker" = "darkgreen")) +
  theme_bw()
```

# Summary

| Concept | Detail |
|--------------------------------------|----------------------------------|
| **Model** | $Y \mid \beta, \tau \sim N_n(X\beta, \tau^{-1}I_n)$ |
| **Independence prior** | $p(\beta, \tau) = p(\beta)p(\tau)$ |
| **Prior on** $\beta$ | Elicited via conditional means $\tilde{m}_i = \tilde{x}_i'\beta$; induced via $\beta = \tilde{X}^{-1}\tilde{m}$ |
| **Prior on** $\tau$ | $\mathrm{Gamma}(a, b)$ with mode and percentile constraints |
| **Key finding** | $\beta_4 < 0$ suggests smoking reduces the rate of FEV growth with age. Illustrated FEV distribution for 18 year old smoker vs non-smoker. |
