---
title: "Lesson 0: Getting Started"
format: html
---

## Welcome

This lesson introduces the tools we will use throughout the course: Quarto, Git, and NIMBLE.

## Quarto Resources

[Quarto](https://quarto.org/) is a scientific publishing system that lets you combine code, text, and output into reproducible documents.

- **Getting started with Quarto in RStudio**: <https://quarto.org/docs/get-started/hello/rstudio.html>
- **Quarto for Scientists**: <https://qmd4sci.njtierney.com/>

## Git

[Git](https://git-scm.com/) is a version control system that helps you track changes to your code and collaborate with others.

- **Git - the simple guide**: <https://rogerdudler.github.io/git-guide/>


## Installing NIMBLE

[NIMBLE](https://r-nimble.org/) is an R package for Bayesian statistical modeling. Install it by running the following code in R:

```{r}
#| eval: false
install.packages("nimble")
```

After installation, load the package with:

```{r}
#| eval: false
library(nimble)
```

## Github (optional)

You should be able to use your UNM affiliation to get the [Github Student Developer Pack](https://github.com/education/students) for free. That should give you access to [Github Copilot AI](https://github.com/features/copilot).

Here are steps to connect Github to Rstudio.
```{r}
#### 1. Sign up at GitHub.com ################################################

## If you do not have a GitHub account, sign up here:
## https://github.com/join

# ----------------------------------------------------------------------------

#### 2. Install git ##########################################################

## If you do not have git installed, please do so: 
## Windows ->  https://git-scm.com/download/win
## Mac     ->  https://git-scm.com/download/mac
## Linux   ->  https://git-scm.com/download/linux
##             or: $ sudo dnf install git-all

# ----------------------------------------------------------------------------

### 3. Configure git with Rstudio ############################################

## set your user name and email (change these to your name and address)
usethis::use_git_config(user.name = "Miheer Dewaskar", user.email = "miheerdew@gmail.com")

## create a personal access token for authentication:
usethis::create_github_token() 
## in case usethis version < 2.0.0: usethis::browse_github_token() (or even better: update usethis!)

## Note for Linux users:
## credentials::set_github_pat() (in line 34) might store your PAT in a memory cache that
## expires after 15 minutes or when the computer is rebooted. You thus may wish to do 
## extend the cache timeout to match the PAT validity period:
usethis::use_git_config(credential.helper="cache --timeout=2600000") #> cache timeout ~30 days

## set personal access token:
credentials::set_github_pat("YourPAT")

## or store it manually in '.Renviron':
usethis::edit_r_environ()
## store your personal access token in the file that opens in your editor with:
## GITHUB_PAT=xxxyyyzzz
## and make sure '.Renviron' ends with a newline

# ----------------------------------------------------------------------------

#### 4. Restart R! ###########################################################

# ----------------------------------------------------------------------------

#### 5. Verify settings ######################################################

usethis::git_sitrep()

## Your username and email should be stated correctly in the output. 
## Also, the report shoud cotain something like:
## 'Personal access token: '<found in env var>''

## If you are still having troubles, read the output carefully.
## It might be that the PAT is still not updated in your `.Renviron` file.
## Call `usethis::edit_r_environ()` to update that file manually.

# ----------------------------------------------------------------------------

## THAT'S IT!
```